When easy is not preferred: A discounting paradigm to assess load-independent task preference

### 2023-03-13 06_Paper/COG-ED/Stage 2/Manuscript.Rmd
continued comments

The sample of the pilot study consisted of $N=15$ participants (53.3% female, $M=24.43$ ($SD=3.59$) years old).
One participant's data was removed because they misunderstood the instruction.
Due to a technical error the subjective task load data of one participant was incomplete, so the hypotheses involving the NASA Task Load Index <!-- CK: TLX --> were analyzed with $n=14$ data sets.
The results showed increases in subjective and objective task load measures with higher $n$-back level.
Importantly, SVs were lower for higher $n$-back levels, but not different between 1- and 2-back, which shows that the easiest level is not universally preferred.
The LMM <!-- CK: Konsitenz: LMM oder MLM -->revealed $n$-back level as a reliable predictor of SV, even after controlling for declining task performance ($d'$ and median RT).
NASA-TLX scores were higher with higher $n$, and lower for the group with lower NFC scores, but NFC and $n$-back level did not interact.
All results are detailed in the Supplementary Material.

@@

When asking participants what motivated their decisions in the effort discounting, `r format(round((table(data_quest$motivation)[1]/sum(table(data_quest$motivation)))*100, digits = 1), nsmall = 1)`% <!-- CK: zwischen Zahl und % braucht es ein Leerzeichen -->stated that they wanted to avoid boredom, `r format(round((table(data_quest$motivation)[2]/sum(table(data_quest$motivation)))*100, digits = 1), nsmall = 1)`% stated that they wanted a challenge, `r format(round((table(data_quest$motivation)[3]/sum(table(data_quest$motivation)))*100, digits = 1), nsmall = 1)`% stated that they wanted to avoid effort, and `r format(round((table(data_quest$motivation)[4]/sum(table(data_quest$motivation)))*100, digits = 1), nsmall = 1)`% stated that they wanted to relax.
The remaining `r format(round((table(data_quest$motivation)[5]/sum(table(data_quest$motivation)))*100, digits = 1), nsmall = 1)`% of participants used the free text field and provided reasons such as "I wanted a fair relation of effort and reward.", "I wanted the fun that I had in the more challenging levels.", "I wanted to maximize reward first and minimize effort second.", or "I did not want to perform poorly when I was being paid for it.".
Figure S.1 in the Supplementary Material shows the different motivations in the context of the SVs per $n$-back level.

@@

Consequently, we had to adapt the multi level model <!-- CK: "MLM" --> to incorporate this non-linear trend.
To apply the contrast to the $n$-back levels, we had to turn the variables into a factor, with two consequences:
Centered variables cannot be turned into factors, so we entered the variable level in its raw form, and factors cannot be used as random slopes, so the model is now defined as:
$$
SV \sim level\ + d' + median RT + (1|subject)
$$
This means that the intercept still varies between subjects, but there are no random slopes anymore.
To provide more than one observation per factor level, we used the two rounds per $n$-back level per subject, rather than $n$-back levels per subject.
<!-- CK: ok, ich sehe gerade, dass hier die Änderungen erklärt werden. -->
The ICC of the null model indicated that there is a correlation of $r=$ `r sub("^(-?)0.", "\\1.", sprintf("%.2f", round(model0_h2b_ICC, digits = 3)))` between the SVs of a subject, i.e. that `r paste0(formatC(model0_h2b_ICC * 100, format = "f", digits = 2), "%")` of variance in SVs can be explained by differences between participants.
We did not use an optimization algorithm to improve the fit of the random intercept model. <!-- CK: ist das ncoh so und warum nicht? -->
A total of `r nrow(excl_cases_h2b)` data points from `r length(unique(excl_cases_h2b$subject))` participants were excluded, because the residuals exceeded 3 SD above the mean. <!-- CK: wie gesagt, das sollte noch angepasst bzw. rausgenommen werden. Vielleicht könnten wir so etwas schreiben wie: "We did not perform model criticism because we now believe that model-compliant exclusion of data is not a good scientific practice." --> 
The results of the final model are displayed in Table\ \@ref(tab:H2b-results).

@@

To investigate the dependency of the model results on the RT preprocessing, we conducted a specification curve analysis (Figure\ \@ref(fig:sca-plot)). <!-- CK: ich dachte, das wäre raus und die SCA wäre nur für H3? -->

@@

The rmANOVA on the association between NFC scores and NASA-TLX scores revealed a main effect of $n$-back level (`r apa_print(hypothesis3b_model)$full_result$level`, `r apa_print(hypothesis3b_BF_levels)$statistic$level`) and an interaction between $n$-back level and NFC scores (`r apa_print(hypothesis3b_model)$full_result$nfcmedian_level`), but no main effect of NFC scores (`r apa_print(hypothesis3b_model)$full_result$nfcmedian`, `r apa_print(hypothesis3b_BF_levels)$statistic$nfcmedian`).
Post-hoc tests showed that the participants with NFC scores below the median had higher NASA-TLX <!-- CK: Konsistenz: NASA-TLX oder NASA TLX --> scores for 3-back (`r apa_print(pairs(hypothesis3b_emm_interact))$statistic$X3_High_low`, `r apa_print(BayesFactor::ttestBF(x = h3b_data$ntlx[h3b_data$nfcmedian == "low" & h3b_data$level == 3], y = h3b_data$ntlx[h3b_data$nfcmedian == "high" & h3b_data$level == 3], progress = FALSE, paired = FALSE))$statistic`) and for 4-back (`r apa_print(pairs(hypothesis3b_emm_interact))$statistic$X4_High_low`, `r apa_print(BayesFactor::ttestBF(x = h3b_data$ntlx[h3b_data$nfcmedian == "low" & h3b_data$level == 4], y = h3b_data$ntlx[h3b_data$nfcmedian == "high" & h3b_data$level == 4], progress = FALSE, paired = FALSE))$statistic`) than those with NFC scores above the median.
Regardless of NFC scores, NASA-TLX scores were higher for the more difficult level in each pair of $n$-back levels (Figure\ \@ref(fig:h3b-plot)).

@@

This Registered Report aimed to adapt the Cognitive Effort Discounting (COG-ED) paradigm by Westbrook et al.[-@Westbrook2013], which can estimate <!-- CK: which estimates or estimating --> subjective values of different $n$-back levels, into the Cognitive and Affective Discounting (CAD) paradigm, which can estimate <!-- CK: s.o. --> SVs of tasks without defaulting to the objective task load as a benchmark.
For this purpose, we changed the way in which the discounting options are presented to the participants, based the anchor on their own choices, and computed SVs across multiple combinations of task levels.
The analyses were closely aligned with those in Westbrook et al.[-@Westbrook2013] to demonstrate the changes in SVs brought about by the new paradigm.
This study also applied the CAD paradigm to an emotion regulation task, the results of which are detailed in a second Registered Report[@ScheffelZerna2022].

@@

The performance measure $d'$ did not differ across $n$-back levels, but the RT increased from 1- to 2- to 3-back and then remained on a high level for 4-back.
This points to three important characteristics of the $n$-back task in this context.
Firstly, RT as a valid group-level indicator of performance might only be useful for levels up to $n=3$, and could be used to investigate inter-individual differences for $n>3$. <!-- CK: Warum für n größer 3? -->
Secondly, there is a speed-accuracy tradeoff in the first three levels, that might even re-emerge in higher levels, where $d'$ would decline and RT would remain stable.
And lastly, the fact that neither accuracy nor speed is an informative performance measure by itself has been observed before[@Meule2017] and both show different associations with various measures of intelligence[@Jaeggi2010], suggesting that they should always be reported as separate indices.
Additionally, $d'$ might not have differed across $n$-back levels because the manipulation of task load is not strictly continuous.
Several participants said that they perceived 3-back as more difficult than 4-back because they found it is easier to remember chunks of stimuli when $n$ was an even number than when $n$ was an odd number.

All NASA-TLX subscales differed across $n$-back levels, but the effort and mental load subscales were the only ones to consistently increase across all levels.
This would support the notion of the $n$-back task offering a continuous manipulation of task load, as <!-- CK: "at"? --> least subjectively.
Ratings on the frustration and time subscales increased and ratings on the performance subscale decreased until 3-back and then remained stable.
This pattern is akin to the RT, which also increased and then remained stable.
Ratings on the physical load subscale increased with $n$-back levels, but not between 2- and 3-back and 3- and 4-back, respectively.

@@

When developing a new paradigm, it is challenging to decide on the optimal analysis strategy, as every hypothesis is based on expected data patterns rather than previous findings.
While the Stage 1 review process made the analyses as robust as possible, there were still unknown factors that should be addressed by future studies.
For instance, the differences between participants with higher and lower NFC should be investigated with extreme groups rather than a median split, especially in academic samples where NFC can be expected to be higher on average and more narrow in range.
Additionally, we expected the SVs of participants with lower NFC scores to peak at 1-back and the SVs of those with higher scores to peak at 2-back, but the way the SVs of both groups appeared to drift apart in the higher $n$-back levels suggests that an analysis of those levels would be more fruitful in determining group differences.
Future studies could create a stronger separation between the concepts investigated <!-- CK: welche Konzepte sind das? -->in this study, and model the SVs and their task-related influencing factors first, before looking at (non-linear) associations with personality.
Another important point is the instruction, not just for the $n$-back task, but for the effort discounting as well.
We had to exclude several participants for misunderstanding the task instruction, so we will add a visual instruction or a training next time.
And even though the participants were instructed to do the effort discounting with the aim to be satisfied with their choices instead of trying to increase the rewards, we cannot be sure that they did so.
One might also argue that the 2€ reward range was not large enough to be an incentive for effort expenditure.
However, findings by Bialaszek et al.[-@Bialaszek2017] suggest that participants are actually more sensitive to effort when the reward is small.
Nevertheless, we exceeded the largest required sample size by `r round(nrow(data_quest)/53, digits = 1)` times, which gives our analyses high statistical power, and we adhered to the agreed upon analyses of Stage 1.

@@

Effort and reward are two highly subjective concepts with relevance in daily life.
With each decision an individual makes, they must weigh the required effort against the expected reward to decide if and how to behave in that situation.
So far, effort discounting paradigms have relied on the assumption that the task that is objectively easiest is the one that is preferred by everyone, and each more difficult task is simply being devalued compared to the easy one.
However, effort-related traits such as Need for Cognition suggest that this is not the case.
Therefore, we developed a paradigm that allows effort discounting independent of objective task load, which we tested using an $n$-back task.
The results showed that many participants <!-- CK: 34.5 % würde ich nicht als "many" verstehen --> indeed preferred a more or even the most difficult $n$-back level.
Spanning the entire sample, these preferences took the shape of a declining logistic curve across $n$-back levels.
While the subjective value declined with increasing levels, it increased with better performance as measured in $d'$, and was unaffected by the reaction time.
Participants with Need for Cognition scores above the median reported lower subjective task load in and less aversion to more difficult levels.
However, they did not have higher subjective values per se, which was likely due to our choice of median split and our assumption that these group differences would emerge in lower levels.
In fact, the reaction time and self-report data suggest that individual differences emerge especially from 3-back upwards, emphasizing the need for tasks with high discriminatory power and effort discounting paradigms with flexible, participant-centered mechanisms.
The CAD paradigm offers this flexibility, and we encourage future studies to question traditional assumptions in the field of effort discounting in the light of these findings, and to re-use this data set for exploratory analyses.

>>>>

The sample of the pilot study consisted of $N=15$ participants (53.3% female, $M=24.43$ ($SD=3.59$) years old).
One participant's data was removed because they misunderstood the instruction.
Due to a technical error the subjective task load data of one participant was incomplete, so the hypotheses involving the NASA-TLX were analyzed with $n=14$ data sets.
The results showed increases in subjective and objective task load measures with higher $n$-back level.
Importantly, SVs were lower for higher $n$-back levels, but not different between 1- and 2-back, which shows that the easiest level is not universally preferred.
The MLM revealed $n$-back level as a reliable predictor of SV, even after controlling for declining task performance ($d'$ and median RT).
NASA-TLX scores were higher with higher $n$, and lower for the group with lower NFC scores, but NFC and $n$-back level did not interact.
All results are detailed in the Supplementary Material.

@@

When asking participants what motivated their decisions in the effort discounting, `r format(round((table(data_quest$motivation)[1]/sum(table(data_quest$motivation)))*100, digits = 1), nsmall = 1)`% stated that they wanted to avoid boredom, `r format(round((table(data_quest$motivation)[2]/sum(table(data_quest$motivation)))*100, digits = 1), nsmall = 1)`% stated that they wanted a challenge, `r format(round((table(data_quest$motivation)[3]/sum(table(data_quest$motivation)))*100, digits = 1), nsmall = 1)`% stated that they wanted to avoid effort, and `r format(round((table(data_quest$motivation)[4]/sum(table(data_quest$motivation)))*100, digits = 1), nsmall = 1)`% stated that they wanted to relax.
The remaining `r format(round((table(data_quest$motivation)[5]/sum(table(data_quest$motivation)))*100, digits = 1), nsmall = 1)`% of participants used the free text field and provided reasons such as "I wanted a fair relation of effort and reward.", "I wanted the fun that I had in the more challenging levels.", "I wanted to maximize reward first and minimize effort second.", or "I did not want to perform poorly when I was being paid for it.".
Figure S.1 in the Supplementary Material shows the different motivations in the context of the SVs per $n$-back level.

@@

Consequently, we had to adapt the MLM to incorporate this non-linear trend.
To apply the contrast to the $n$-back levels, we had to turn the variables into a factor, with two consequences:
Centered variables cannot be turned into factors, so we entered the variable level in its raw form, and factors cannot be used as random slopes, so the model is now defined as:
$$
SV \sim level\ + d' + median RT + (1|subject)
$$
This means that the intercept still varies between subjects, but there are no random slopes anymore.
To provide more than one observation per factor level, we used the two rounds per $n$-back level per subject, rather than $n$-back levels per subject.
The ICC of the null model indicated that there is a correlation of $r=$ `r sub("^(-?)0.", "\\1.", sprintf("%.2f", round(model0_h2b_ICC, digits = 3)))` between the SVs of a subject, i.e. that `r paste0(formatC(model0_h2b_ICC * 100, format = "f", digits = 2), "%")` of variance in SVs can be explained by differences between participants.
We did not use an optimization algorithm to improve the fit of the random intercept model.
We did not exclude any data points.
The results of the final model are displayed in Table\ \@ref(tab:H2b-results).

@@

To investigate the dependency of the model results on the RT preprocessing, we conducted a specification curve analysis (Figure\ \@ref(fig:sca-plot)).

@@

The rmANOVA on the association between NFC scores and NASA-TLX scores revealed a main effect of $n$-back level (`r apa_print(hypothesis3b_model)$full_result$level`, `r apa_print(hypothesis3b_BF_levels)$statistic$level`) and an interaction between $n$-back level and NFC scores (`r apa_print(hypothesis3b_model)$full_result$nfcmedian_level`), but no main effect of NFC scores (`r apa_print(hypothesis3b_model)$full_result$nfcmedian`, `r apa_print(hypothesis3b_BF_levels)$statistic$nfcmedian`).
Post-hoc tests showed that the participants with NFC scores below the median had higher NASA-TLX scores for 3-back (`r apa_print(pairs(hypothesis3b_emm_interact))$statistic$X3_High_low`, `r apa_print(BayesFactor::ttestBF(x = h3b_data$ntlx[h3b_data$nfcmedian == "low" & h3b_data$level == 3], y = h3b_data$ntlx[h3b_data$nfcmedian == "high" & h3b_data$level == 3], progress = FALSE, paired = FALSE))$statistic`) and for 4-back (`r apa_print(pairs(hypothesis3b_emm_interact))$statistic$X4_High_low`, `r apa_print(BayesFactor::ttestBF(x = h3b_data$ntlx[h3b_data$nfcmedian == "low" & h3b_data$level == 4], y = h3b_data$ntlx[h3b_data$nfcmedian == "high" & h3b_data$level == 4], progress = FALSE, paired = FALSE))$statistic`) than those with NFC scores above the median.
Regardless of NFC scores, NASA-TLX scores were higher for the more difficult level in each pair of $n$-back levels (Figure\ \@ref(fig:h3b-plot)).

@@

This Registered Report aimed to adapt the Cognitive Effort Discounting (COG-ED) paradigm by Westbrook et al.[-@Westbrook2013], which estimates subjective values of different $n$-back levels, into the Cognitive and Affective Discounting (CAD) paradigm, which estimates SVs of tasks without defaulting to the objective task load as a benchmark.
For this purpose, we changed the way in which the discounting options are presented to the participants, based the anchor on their own choices, and computed SVs across multiple combinations of task levels.
The analyses were closely aligned with those in Westbrook et al.[-@Westbrook2013] to demonstrate the changes in SVs brought about by the new paradigm.
This study also applied the CAD paradigm to an emotion regulation task, the results of which are detailed in a second Registered Report[@ScheffelZerna2022].

@@

The performance measure $d'$ did not differ across $n$-back levels, but the RT increased from 1- to 2- to 3-back and then remained on a high level for 4-back.
This points to three important characteristics of the $n$-back task in this context.
Firstly, RT as a valid group-level indicator of performance might only be useful for levels up to $n=3$, and could be used to investigate inter-individual differences for $n>3$.
Secondly, there is a speed-accuracy tradeoff in the first three levels, that might even re-emerge in higher levels, where $d'$ would decline and RT would remain stable.
And lastly, the fact that neither accuracy nor speed is an informative performance measure by itself has been observed before[@Meule2017] and both show different associations with various measures of intelligence[@Jaeggi2010], suggesting that they should always be reported as separate indices.
Additionally, $d'$ might not have differed across $n$-back levels because the manipulation of task load is not strictly continuous.
Several participants said that they perceived 3-back as more difficult than 4-back because they found it is easier to remember chunks of stimuli when $n$ was an even number than when $n$ was an odd number.

All NASA-TLX subscales differed across $n$-back levels, but the effort and mental load subscales were the only ones to consistently increase across all levels.
This would support the notion of the $n$-back task offering a continuous manipulation of task load, at least subjectively.
Ratings on the frustration and time subscales increased and ratings on the performance subscale decreased until 3-back and then remained stable.
This pattern is akin to the RT, which also increased and then remained stable.
Ratings on the physical load subscale increased with $n$-back levels, but not between 2- and 3-back and 3- and 4-back, respectively.

@@

When developing a new paradigm, it is challenging to decide on the optimal analysis strategy, as every hypothesis is based on expected data patterns rather than previous findings.
While the Stage 1 review process made the analyses as robust as possible, there were still unknown factors that should be addressed by future studies.
For instance, the differences between participants with higher and lower NFC should be investigated with extreme groups rather than a median split, especially in academic samples where NFC can be expected to be higher on average and more narrow in range.
Additionally, we expected the SVs of participants with lower NFC scores to peak at 1-back and the SVs of those with higher scores to peak at 2-back, but the way the SVs of both groups appeared to drift apart in the higher $n$-back levels suggests that an analysis of those levels would be more fruitful in determining group differences.
Future studies could create a stronger separation between the concepts investigated in this study (discounting curve, effort perception, performance, SV computation, NFC), and model the SVs and their task-related influencing factors first, before looking at (non-linear) associations with personality.
Another important point is the instruction, not just for the $n$-back task, but for the effort discounting as well.
We had to exclude several participants for misunderstanding the task instruction, so we will add a visual instruction or a training next time.
And even though the participants were instructed to do the effort discounting with the aim to be satisfied with their choices instead of trying to increase the rewards, we cannot be sure that they did so.
One might also argue that the 2€ reward range was not large enough to be an incentive for effort expenditure.
However, findings by Bialaszek et al.[-@Bialaszek2017] suggest that participants are actually more sensitive to effort when the reward is small.
Nevertheless, we exceeded the largest required sample size by `r round(nrow(data_quest)/53, digits = 1)` times, which gives our analyses high statistical power, and we adhered to the agreed upon analyses of Stage 1.

@@

Effort and reward are two highly subjective concepts with relevance in daily life.
With each decision an individual makes, they must weigh the required effort against the expected reward to decide if and how to behave in that situation.
So far, effort discounting paradigms have relied on the assumption that the task that is objectively easiest is the one that is preferred by everyone, and each more difficult task is simply being devalued compared to the easy one.
However, effort-related traits such as Need for Cognition suggest that this is not the case.
Therefore, we developed a paradigm that allows effort discounting independent of objective task load, which we tested using an $n$-back task.
The results showed that many participants indeed preferred a more or even the most difficult $n$-back level.
Spanning the entire sample, these preferences took the shape of a declining logistic curve across $n$-back levels.
While the subjective value declined with increasing levels, it increased with better performance as measured in $d'$, and was unaffected by the reaction time.
Participants with Need for Cognition scores above the median reported lower subjective task load in and less aversion to more difficult levels.
However, they did not have higher subjective values per se, which was likely due to our choice of median split and our assumption that these group differences would emerge in lower levels.
In fact, the reaction time and self-report data suggest that individual differences emerge especially from 3-back upwards, emphasizing the need for tasks with high discriminatory power and effort discounting paradigms with flexible, participant-centered mechanisms.
The CAD paradigm offers this flexibility, and we encourage future studies to question traditional assumptions in the field of effort discounting in the light of these findings, and to re-use this data set for exploratory analyses.
